Extending competitive accessibility for multi-modal analysis

### 2023-07-30 Multi-mode-acc-methods.qmd
Presentation file commit pt.2. and draft manuscript documents.

---
title: "Extending competitive accessibility for multi-modal analysis"
author: Anastasia Soukhov, Javier Tarriño Ortiz, Julio A. Soria-Lara, Antonio Páez
format: docx
bibliography: bibliography.bib
---
```{r load-packages, include=FALSE, cache=FALSE}
# library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax

@@

ANother example of proportional allocation, BFCA

<!-- policy applications of accessibility broadly, but what's missing is multi-modal accessibility methods -->
The consideration of both supply of opportunities and demand from the population is important for policy applications. Spatial availability can be used to create supply per capita benchmarks for locations, and the delta of supply from this benchmark to existing conditions can be used to motivate policy. Currently, conventional _unconstrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons. Furthermore, exisiting competitive measures are difficult to interpret.

However, this is a gap in the literature. Researchers and transport planners needs accessibility tools that interpretable consider competition. This is the aim of this paper. We plan to expland to this measure to multi-modal systems. opportunities are avialble to users of different modes. The cost of reaching them is not uniform.

<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility. Take for instance, measuring the accessibility to opportunities for a region for car and transit modes. The literature suggests that these modes are in direct competition. A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other. Thus, the potential for the population to interact with opportunities is directly dependant on ... the opportunities..  -->

@@

# Illustrative example: using spatial availability for multiple modes

@@

The general form of spatial availability is a sum: this is similar to Hansen-type accessibility (popular _unconstrained_ measure) and Shen-type accessibility (popular _competitive_ and _unconstrained_ measure). What makes spatial availability a _constrained_ measure are the balancing factors that allocate opportunities proportionally. The balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).

@@

However, how do the populations access these jobs? In this synthetic example, the population either uses the regular mode $z$ or uses the faster mode $x$. Their choice in mode is already assigned: the mode selected can be seen as a characteristic of the population, or a way to subset the total population. In this way, the introduction of this population-characteristic of mode selection adds another layer of competition to the mass effect. We demonstrated that employment center 1 allocates $23,809.52$ jobs to $A$ but 20% of these jobs are allocated to population using $z$ ($10,000$ of the $50,000$ population in $A$) and 80% is allocated to the population using the faster mode $x$ ($40,000$ of the $50,000$ population in $A$). Similarly, employment centers 2 and 3 allocate $71,428.57$ jobs to $B$ (the urban city) of which 66% are allocated to the population using $z$ and 33% to the population using $x$ mode and the $7,142.857$ jobs allocated to $C$ (the satellite city) is split between 30% of the population using mode $z$ and 70% of the population using mode $x$.

@@

#NOTE: maybe we can try a distance based distribution???? Not really convinced why would could - perhaps to include all modes?

@@

Estimate f value using these averaged travel times:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
OD_2 <- OD_2 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_gamma$estimate["shape"], 
                                                  HW_transit_gamma$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])))

#also for zn208
OD_3 <- OD_3 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_gamma$estimate["shape"], 
                                                  HW_transit_gamma$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])))
```

@@

#zn208
SA_im_zn208 <- OD_SA_zn208 %>%
  group_by(VORI_CD_ZT208, 
           mode) %>%
  summarize(V_im = sum(SA_ij),
            .groups = "drop")

@@

@@

SA_im_zn208 <- SA_im_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()
```

@@

@@

Plot V (spatial availability) by origin-mode (first zn1259 then zn208):
```{r}
SA_im_V_plot <- ggplot() +
  geom_sf(data = SA_im,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(V["im"]),
                       na.value = "grey90",
                       direction = 1)+
  geom_sf(data = Madrid_muni_border, # border for Madrid munim
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)

#zn208
SA_im_V_zn208_plot <- ggplot() +
  geom_sf(data = SA_im_zn208,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(V["im"]),
                       na.value = "grey90",
                       direction = 1,
                       #trans = "log",
                       #breaks=c(0, 10, 100, 1000, 10000)
                       )+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
```
```{r}
SA_im_V_plot
ggsave("Presentations/SA_im_V_zn1259_plot.png")
```
```{r}
SA_im_V_zn208_plot
ggsave("Presentations/SA_im_V_zn208_plot.png")
```

>>>>

---
title: "Extending competitive accessibility for multi-modal analysis"
author: Anastasia Soukhov, Javier Tarriño Ortiz, Julio A. Soria-Lara, Antonio Páez
format: docx
bibliography: mybibfile.bib
---
```{r load-packages, include=FALSE, cache=FALSE}
# library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax

@@

Another example of proportional allocation, BFCA

<!-- policy applications of accessibility broadly, but what's missing is multi-modal accessibility methods -->
The consideration of both supply of opportunities and demand from the population is important for policy applications. Spatial availability can be used to create supply per capita benchmarks for locations, and the delta of supply from this benchmark to existing conditions can be used to motivate policy. Currently, conventional _unconstrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons. Furthermore, existing competitive measures are difficult to interpret.

However, this is a gap in the literature. Researchers and transport planners needs accessibility tools that interpretable consider competition. This is the aim of this paper. We plan to expand to this measure to multi-modal systems. opportunities are available to users of different modes. The cost of reaching them is not uniform.

@@

# Illustrative example: operationalizing spatial availability for multiple modes

@@

The general form of spatial availability is a sum: this is similar to Hansen-type accessibility (popular _unconstrained_ measure) and Shen-type accessibility (popular _competitive_ and _unconstrained_ measure). What makes spatial availability both competitive and constrained are the balancing factors that allocate opportunities proportionally. The balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect). Spatial availability considers demand for opportunities based on the magnitude of demand for opportunities and the travel-cost to reach those opportunities (i.e., competition) as well as preserving the sum of opportunities (i.e., constrained) such that scores are a proportion of the total number of opportunities in the region.

<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility. Take for instance, measuring the accessibility to opportunities for a region for car and transit modes. The literature suggests that these modes are in direct competition. A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other. Thus, the potential for the population to interact with opportunities is directly dependant on the opportunities t  -->

See @soukhovIntroducingSpatialAvailability2023 for details, but in short, the proportional allocation mechanism of $F^t_{ij}$ ensures that each $V_i$ value assigned to each $i$ sums to the total number of opportunities. This is important, because opportunities in a region are finite. If one person interacts with the opportunity at a given time, it is taken. In this way, if one person has the ability to reach more opportunities through a faster mode, for instance, than that person has more opportunities they can potentially interact with than other people. In a region with finite opportunities, one person being able to access more opportunities means there are fewer opportunities available for people using slower modes.

can be formulated to consider  characteristics of different population groups. For instance, consider two sub-populations at population center $A$. One sub-population uses mode $x$ and the other uses mode $z$ to access jobs at employment center $1$. Mode $x$ is faster than mode $z$, so the sub-population that use $x$ can potentially access more jobs at employment center $1$ than the other sub-population as a result of the lower travel cost. Furthermore, the sub-population using the faster mode $x$ can potentially access more jobs at further employment centers (say $2$ and $3$) than the sub-population using the slower mode. This puts the sub-population using the faster (lower travel cost) mode $x$ at an accessibility advantage. If that sub-population is also within a population center that is low-population (relative to the region) and central located to employment centers (relative to the region), then this sub-population using the faster mode $x$ is the most advantaged of all sub-populations using any mode or located at any population center.

The illustrative example described is represented in Figure \ref{fig:Fig1} and was inspired by the single-mode example used in @shen1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023. Population center $A$ can be seen as a Suburban population center; it is closest to its own relatively large employment center at $A$, close to the Urban area's equally large employment center $B$, and has a population that is smaller than the Urban center $2$ and larger than the rural center $3$. Consider the number of jobs to be finite in the region. From the perspective of accessibility, the sub-population that is in a center that is the least populated, is the most centrally located to largest amount of jobs, is located away from densely populated centers, and uses the lowest travel cost mode $x$ can potentially access the most job opportunities. From the opposite perspective, sub-populations using the slower mode $z$ and especially those located in population centers that are not centrally located to opportunities and/or close to highly dense population centers, are at an access disadvantage. The number of job opportunities are finite, and the sub-populations using faster modes and/or centrally located take a larger proportion of opportunities, leaving fewer for other populations.

@@

However, how do the populations access these jobs? In this example, the population either uses the regular mode $z$ or uses the faster mode $x$. Their choice in mode is already assigned: the mode selected can be seen as a characteristic of the population, or a way to subset the total population. In this way, the introduction of this population-characteristic of mode selection adds another layer of competition to the mass effect. We demonstrated that employment center 1 allocates $23,809.52$ jobs to $A$ but 20% of these jobs are allocated to population using $z$ ($10,000$ of the $50,000$ population in $A$) and 80% is allocated to the population using the faster mode $x$ ($40,000$ of the $50,000$ population in $A$). Similarly, employment centers 2 and 3 allocate $71,428.57$ jobs to $B$ (the urban city) of which 66% are allocated to the population using $z$ and 33% to the population using $x$ mode and the $7,142.857$ jobs allocated to $C$ (the satellite city) is split between 30% of the population using mode $z$ and 70% of the population using mode $x$.

@@

@@

Estimate f value using these averaged travel times:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
OD_2 <- OD_2 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_gamma$estimate["shape"], 
                                                  HW_transit_gamma$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])),
         jobs_f = jobs*f) #for Hansen-type accessibility (unconstrained)

#also for zn208
OD_3 <- OD_3 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_gamma$estimate["shape"], 
                                                  HW_transit_gamma$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])),
         jobs_f = jobs*f) #for Hansen-type accessibility (unconstrained)
```

## Calculating hansen-type accessibility (for comparision)
```{r, echo=FALSE, warning=FALSE, message=FALSE}
OD_Si <- OD_2 %>%
  group_by(VORIZT1259, mode) %>%
  summarize(S_i = sum(jobs_f)) # Hansen-type accessibility (unconstrained)

#also for zn208
OD_Si_zn208 <- OD_3 %>%
  group_by(VORI_CD_ZT208, mode) %>%
  summarize(S_i = sum(jobs_f)) # Hansen-type accessibility (unconstrained)
```

@@

#zn208
SA_im_zn208 <- OD_SA_zn208 %>%
  group_by(VORI_CD_ZT208, 
           mode) %>%
  summarize(V_im = sum(SA_ij),
            population_m = mean(workers),
            .groups = "drop")

@@

OD_Si <- OD_Si %>%
  left_join(zn1259 %>%
              dplyr::select(ZT1259),
            by = c("VORIZT1259" = "ZT1259")) %>%
  st_as_sf()

@@

SA_im_zn208 <- SA_im_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()

OD_Si_zn208 <- OD_Si_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()
```

@@

```{r}
#mapping population by mode!
ggplot() +
  geom_sf(data = SA_im_zn208,
          aes(fill = population_m),
          color = NA) +
  scale_fill_distiller(palette = "Reds", #legend scale bar
                       na.value = "grey90",
                       direction = 1)+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
ggsave("Presentations/im_populations_zn208_plot.png")
```

@@

Plot V (spatial availability) by origin-mode (first zn1259 then zn208). And for comparision, plot hansen-type accessibility by origin-mode (first zn1259 then zn208):
```{r}
SA_im_V_plot <- ggplot() +
  geom_sf(data = SA_im,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(V["im"]),
                       na.value = "grey90",
                       direction = 1)+
  geom_sf(data = Madrid_muni_border, # border for Madrid munim
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)

#zn208
#the accessibility plot (Hansen-type)
S_im_plot_zn208_plot <- ggplot() +
  geom_sf(data = OD_Si_zn208,
          aes(fill = S_i),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(S["im"]),
                       na.value = "grey90",
                       direction = 1)+
  geom_sf(data = Madrid_muni_border, # border for Madrid munim
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)

SA_im_V_zn208_plot <- ggplot() +
  geom_sf(data = SA_im_zn208,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(V["im"]),
                       na.value = "grey90",
                       direction = 1,
                       #trans = "log",
                       #breaks=c(0, 10, 100, 1000, 10000)
                       )+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
```
```{r}
SA_im_V_plot
ggsave("Presentations/SA_im_V_zn1259_plot.png")
```
```{r}
S_im_plot_zn208_plot
ggsave("Presentations/Hansen_im_V_zn208_plot.png")
```

```{r}
SA_im_V_zn208_plot
ggsave("Presentations/SA_im_V_zn208_plot.png")
```
